---
title: "Fisher's Exact Test"
date: "December 16, 2025"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: pygments
fontsize: 12pt
geometry: margin=1in
urlcolor: blue
header-includes:
  - \usepackage[font=small,labelfont=bf]{caption}
  - \usepackage[default]{lato}
  - \usepackage[scale=0.9]{sourcecodepro}
  - \usepackage[T1]{fontenc}
---

\newpage

# Example: Old vs New Treatment

We are researching a new medical treatment and wish to discover whether it is better than the old one.

```{r}
M <- matrix(c(5,2,2,4),2,2)
dimnames(M) <- list(treatment=c("new","old"),cured=c(TRUE,FALSE))
M
```

```{r}
# x: contingency table (2 by 2)
# alternative: Alternative hypothesis
# Testing if being in 'new' increases the odds of being in 'TRUE'

fisher.test(x = M, alternative = "greater")
```

# How to use 'alternative'

```{r}
M <- matrix(c('a','c','b','d'),2,2)
dimnames(M) <- list(c("Row 1","Row 2"),c("Col 1","Col 2"))
knitr::kable(M)
```

The **Odds Ratio (OR)** is estimated roughly as:
$$
OR \approx \frac{a \times d}{b \times c}
$$

The test compares:

- the odds of an event in **Row 1** falling into **Column 1** versus
- the odds of an event in **Row 2** falling into **Column 1**.

## alternative = "greater"

- **Row 1 is more associated with Column 1** than Row 2 is.
- Testing if the **Odds Ratio > 1**.
- In English: "Being in Row 1 increases the odds of being in Column 1."

## alternative = "less"

- **Row 1 is less associated with Column 1** than Row 2 is.
- Testing if the **Odds Ratio < 1**.
- In English: "Being in Row 1 decreases the odds of being in Column 1."

## alternative = "two.sided"

- Testing if the **Odds Ratio $\neq$ 1**.
- In English: "Row 1 and Row 2 have different distributions across the columns, but I don't assume a specific direction beforehand."
